Weighted Model Counting with Conditional Weights for Bayesian Networks

### 2021-05-21 doc/UAI_paper/paper.tex
Cleanup
Update the UAI paper (still incomplete)

\documentclass{uai2021} % for initial submission
% \documentclass[accepted]{uai2021} % after acceptance, for a revised
                                    % version; also before submission to
                                    % see how the non-anonymous paper
                                    % would look like
%% There is a class option to choose the math font
% \documentclass[mathfont=cm]{uai2021} % Computer Modern math instead of
                                       % ptmx, like default for UAI ≤2020
% \documentclass[mathfont=newtx]{uai2021} % newtx fonts (improves upon
                                          % ptmx; less tested, no support)
% NOTE: Only keep *one* line above as appropriate, as it will be replaced
%       automatically for papers to be published. Do not make any other
%       change above this note for an accepted version.

%% Choose your variant of English; be consistent
% \usepackage[american]{babel}
\usepackage[british]{babel}

%% Some suggested packages, as needed:
\usepackage{natbib} % has a nice set of citation styles and commands
    \bibliographystyle{plainnat}
    \renewcommand{\bibsection}{\subsubsection*{References}}
\usepackage{mathtools} % amsmath with fixes and additions
\usepackage[binary-units]{siunitx} % for proper typesetting of numbers and units
\usepackage{booktabs} % commands to create good-looking tables
\usepackage[table]{xcolor}
\usepackage{tikz} % nice language for creating drawings and diagrams

\usepackage{complexity}
\usepackage[capitalise]{cleveref}
\usepackage{multirow}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{thm-restate}
\usepackage{subcaption}
\usepackage[ruled,vlined]{algorithm2e}

@@

\title{Weighted Model Counting with Conditional Weights for Bayesian Networks}

% The standard author block has changed for UAI 2021 to provide
% more space for long author lists and allow for complex affiliations
%
% All author information is authomatically removed by the class for the
% anonymous submission version of your paper, so you can already add your
% information below.
%
% Add authors in order of decreasing contribution
\author[1]{\href{mailto:Harry Q. Bovik <harryq@example.edu>?Subject=Your UAI 2021 paper}{Harry~Q.~Bovik}{}} % Lead author
\author[2]{Coauthor~One}
\author[1,2]{Coauthor~Two}
\author[3]{Further~Coauthor}
\author[1]{Further~Coauthor}
\author[3]{Further~Coauthor}
\author[3,1]{Further~Coauthor}
% Add affiliations after the authors
\affil[1]{%
    Computer Science Dept.\\
    Cranberry University\\
    Pittsburgh, Pennsylvania, USA
}
\affil[2]{%
    Affiliation\\
    Address\\
    …
}
\affil[3]{…}

@@

\begin{abstract}
  Weighted model counting (WMC) has emerged as the unifying inference mechanism
  across many (probabilistic) domains. Encoding an inference problem as an
  instance of WMC typically necessitates adding extra literals and clauses. This
  is partly so because the predominant definition of WMC assigns weights to
  models based on weights on literals, and this severely restricts what
  probability distributions can be represented. We develop a measure-theoretic
  perspective on WMC and propose a way to encode conditional weights on literals
  analogously to conditional probabilities. This representation can be as
  succinct as standard WMC with weights on literals but can also expand as
  needed to represent probability distributions with less structure. To
  demonstrate the performance benefits of conditional weights over the addition
  of extra literals, we develop a new WMC encoding for Bayesian networks and
  adapt a state-of-the-art WMC algorithm \textsf{ADDMC} to the new format. Our
  experiments show that the new encoding significantly improves the performance
  of the algorithm on most benchmark instances. Furthermore, the same idea can
  be adapted to other WMC algorithms and other problem domains.
\end{abstract}

\section{Introduction}
% TODO: keep mentioning citations along with algorithm and encoding names

@@

@@

\begin{table*}
  \caption{Notation for a logic with two atoms. The elements in both columns are
    listed in the same order.}
  \label{tbl:notation_example}
  \centering
  \begin{tabular}{lcc}
    \toprule
    Name in logic & Boolean-algebraic notation & Set-theoretic notation \\
    \midrule
    Atoms (elements of $U$) & $a, b$ & $a, b$ \\
    \rowcolor{gray!10} Models (elements of $2^U$) & $\neg a \land \neg b, a \land \neg b, \neg a \land b, a \land b$ & $\emptyset, \{a\}, \{b\}, \{a, b\}$ \\
    & $\top$ & $\{ \emptyset, \{a\}, \{b\}, \{a, b\} \}$ \\
    & $\neg a \lor \neg b, a \to b$ & $\{ \emptyset, \{a\}, \{b\} \}, \{ \emptyset, \{a\}, \{a, b\} \}$ \\
    & $b \to a, a \lor b$ & $\{ \emptyset, \{b\}, \{a, b\} \}, \{ \{a\}, \{b\}, \{a, b\} \}$ \\
    & $\neg b, \neg a, a \leftrightarrow b$ & $\{\emptyset, \{a\}\}, \{\emptyset, \{b\}\}, \{\emptyset, \{a, b\}\}$ \\
    & $(a \land \neg b) \lor (b \land \neg a), a, b$ & $\{\{a\}, \{b\}\}, \{\{a\}, \{a, b\}\}, \{\{b\}, \{a, b\}\}$ \\
    & $\neg a \land \neg b, a \land \neg b, \neg a \land b, a \land b$ & $\{\emptyset\}, \{\{a\}\}, \{\{b\}\}, \{\{a, b\}\}$ \\
    \multirow{-7}{*}{Formulas (elements of $2^{2^U}$)} & $\bot$ & $\emptyset$ \\
    \bottomrule
  \end{tabular}
\end{table*}

@@

\subsection{Functions on Boolean Algebras}
% TODO: maybe: it would be easier to talk about pseudo-Boolean functions
% instead, but maybe it's not necessary.

@@

\begin{restatable}{theorem}{measure} \label{prop:measure}
  The function $\mu_\nu$ is a measure.
\end{restatable}

@@

Since many measures of interest may not be factorable, a well-known way to
encode them into instances of WMC is by adding more literals
\citep{DBLP:journals/ai/ChaviraD08}. We can use the measure-theoretic
perspective on WMC to show that this is always possible, however, as ensuing
sections will demonstrate, it can make the inference task much harder in
practice.\footnote{The proofs of this and other theoretical results can be found
  in the appendix.}

\begin{restatable}{theorem}{extendable}
  For any set $U$ and measure $\mu\colon 2^{2^U} \to \mathbb{R}_{\ge 0}$, there
  exists a set $V \supseteq U$, a factorable measure $\mu'\colon 2^{2^V} \to
  \mathbb{R}_{\ge 0}$, and a formula $f \in 2^{2^V}$ such that $\mu(x) = \mu'(x
  \land f)$ for all formulas $x \in 2^{2^U}$.
\end{restatable}

@@

For a binary random variable $X$, $\mathrm{CPT}_X$ is simply a sum of smaller
functions, one for each row of the CPT. If $X$ has more than two values, we also
multiply $\mathrm{CPT}_X$ by `clause' functions that restrict the value of
$\phi(T)$ to zero whenever $|\mathcal{E}(X) \cap T| \ne 1$. For the Bayesian
network in \cref{fig:example_bn}, we get:
\begin{align*}
  \mathrm{CPT_F} &= 0.6[\lambda_{F=1}] \cdot [\lambda_{W=1}] + 0.4[\lambda_{F=0}] \cdot [\lambda_{W=1}] \\
                 &+ 0.1[\lambda_{F=1}] \cdot [\lambda_{W=0}] + 0.9[\lambda_{F=0}] \cdot [\lambda_{W=0}], \\
  \mathrm{CPT_T} &= ([\lambda_{T=l}] + [\lambda_{T=m}] + [\lambda_{T=h}]) \\
                 &\cdot (\overline{[\lambda_{T=l}]} + \overline{[\lambda_{T=m}]}) \cdot (\overline{[\lambda_{T=l}]} + \overline{[\lambda_{T=h}]}) \\
                 &\cdot (\overline{[\lambda_{T=m}]} + \overline{[\lambda_{T=h}]}) \cdot \dots.
\end{align*}

@@

\begin{restatable}{lemma}{cptlemma} \label{lemma:cpt}
  Let $X \in \mathcal{V}$ be a random variable with parents $\mathrm{pa}(X) = \{ Y_1,
  \dots, Y_n \}$. Then $\mathrm{CPT}_X\colon 2^{\mathcal{E}^*(X)} \to
  \mathbb{R}_{\ge 0}$ is such that for any $x \in \im X$ and $(y_1, \dots, y_n)
  \in \prod_{i=1}^n \im Y_i$,
  \[
    \mathrm{CPT}_X (T) = \Pr(X = x \mid Y_1 = y_1, \dots, Y_n = y_n),
  \]
  where $T = \{ \lambda_{X=x} \} \cup \{ \lambda_{Y_i=y_i} \mid i = 1, \dots, n
  \}$.
\end{restatable}

@@

\begin{restatable}{lemma}{fulldistribution} \label{lemma:full_distribution}
  Let $\mathcal{V} = \{X_1, \dots, X_n\}$. Then
  \[
    \phi(T) =
    \begin{cases}
      \Pr(x_1, \dots, x_n) &
      \begin{aligned}
        &\text{if } T = \{ \lambda_{X_i=x_i} \}_{i = 1}^n \text{ for} \\
        &\text{some } \textstyle (x_i)_{i=1}^n \in \prod_{i=1}^n \im X_i
      \end{aligned} \\
      0 & \text{otherwise,}
    \end{cases}
  \]
  for all $T \in 2^U$.
\end{restatable}

We end with \cref{thm:correctness} that shows how $\phi$ can be combined with an
encoding of a single variable-value assignment so that \textsf{ADDMC} would
compute its marginal probability.

\begin{restatable}{theorem}{correctness} \label{thm:correctness}
  For any $X \in \mathcal{V}$ and $x \in \im X$,
  \[
    (\exists_U(\phi \cdot [\lambda_{X=x}]))(\emptyset) = \Pr(X = x).
  \]
\end{restatable}

@@

%\begin{figure*}[t] % TODO: maybe not: describe this in the text
%  \centering
%  \input{melt.tex}%
%  \caption{Average encoding and inference time of each encoding with their
%    original WMC algorithm (on the left) and with \textsf{ADDMC} (on the right)
%    on \emph{solved} instances. The error bars denote \SI{10}{\percent} of
%    standard deviation.}
%  \label{fig:melted}
%\end{figure*}

We compare the six WMC encodings for Bayesian networks when run with both
\textsf{ADDMC} and the WMC algorithms used in the original papers.\footnote{Both
  \texttt{cd05} and \texttt{cd06} cannot be run with most WMC algorithms
  including \textsf{ADDMC} because these encodings allow for additional models
  that the WMC algorithm is supposed to ignore
  \citep{DBLP:conf/ijcai/ChaviraD05,DBLP:conf/sat/ChaviraD06}.} We compare the
encodings with respect to the total time it takes to encode a Bayesian network,
compile it or run a WMC algorithm on it, and extract the (numerical) answer.
Note that while all five papers that introduce other encodings include
experimental comparisons of encoding size, that is not feasible with
\textsf{ADDMC} as even instances that are fully solved in less than
\SI{0.1}{\second} are too big to build the full ADD within reasonable time and
memory limits. The experiments were run on a computing cluster with Intel Xeon
Gold 6138 and Intel Xeon E5-2630 processors\footnote{Each instance is run on the
  same processor for all encodings.} running Scientific~Linux~7 with a
\SI{32}{\gibi\byte} memory limit and a \SI{1000}{\second} timeout on both
encoding and inference. For inference, we use \textsf{Ace} for \texttt{cd05},
\texttt{cd06}, and \texttt{d02};
\textsf{Cachet}\footnote{\url{https://cs.rochester.edu/u/kautz/Cachet/}}
\citep{DBLP:conf/sat/SangBBKP04} for \texttt{sbk05}; and \textsf{c2d}
\citep{DBLP:conf/ecai/Darwiche04} for compilation and \textsf{query-dnnf}
\footnote{\url{http://www.cril.univ-artois.fr/kc/d-DNNF-reasoner.html}} for
answer computation for \texttt{bklm16}. For encoding, we use \textsf{bn2cnf}
\footnote{\url{http://www.cril.univ-artois.fr/KC/bn2cnf.html}} for
\texttt{bklm16}, and \textsf{Ace} for all other encodings (except for
\texttt{cw}, which is implemented in Python).

\textsf{Ace} was not used to encode evidence, as
preliminary experiments revealed that the evidence-encoding implementation
contains bugs that can lead to incorrect answers or a Java exception being
thrown on some instances of the data set (and the source code is not publicly
available). Instead, we simply list all the evidence as additional clauses in
the encoding. Furthermore, to ensure that \texttt{bklm16} (whether run with
\textsf{ADDMC} or \textsf{c2d}) returns correct answers on most instances, we
had to disable one of the improvements that \texttt{bklm16} brings over
\texttt{cd06}, namely, the construction of a scaling factor that `absorbs' one
probability from each CDT \citep{DBLP:conf/ecai/BartKLM16}. For realistic
benchmark instances, this scaling factor can easily be below $10^{-30}$, and
thus would require arbitrary-precision floating-point arithmetic to be usable.
Even a toy Bayesian network with seven binary independent variables with
probabilities $0.1$ and $0.9$ is enough for \textsf{bn2cnf} to output precisely
zero as the scaling factor. We note that this issue likely remained unnoticed
because \citet{DBLP:conf/ecai/BartKLM16} did not attempt to compute numerical
answers in their experiments.

@@

For experimental data, we use the Bayesian networks
available with \textsf{Ace} and \textsf{Cachet}, most of which happen to be
binary. We classify them into the following seven categories:
\begin{itemize*}
\item DQMR and
\item Grid networks as described by \citet{DBLP:conf/aaai/SangBK05},
\item Mastermind, and
\item Random Blocks from the work of \citet{DBLP:journals/ijar/ChaviraDJ06},
\item remaining binary Bayesian networks that include Plan Recognition
  \citep{DBLP:conf/aaai/SangBK05}, Friends and Smokers, Students and Professors
  \citep{DBLP:journals/ijar/ChaviraDJ06}, and \texttt{tcc4f}, and
\item non-binary classic Bayesian networks (\texttt{alarm}, \texttt{diabetes},
  \texttt{hailfinder}, \texttt{mildew}, \texttt{munin1}--\texttt{4},
  \texttt{pathfinder}, \texttt{pigs}, \texttt{water}).
\end{itemize*}
We run \textsf{ADDMC} with each of the five encodings once on each Bayesian
network. % TODO: this should be updated

%Specifically, \texttt{cw} solves the same number of
%instances in \SI{8.144}{\second} as \texttt{d02} does in \SI{1000}{\second},
%which is the second-best encoding in our experiments.

\Cref{fig:cumulative} shows that \texttt{cd05} and
\texttt{cd06} (when run with \textsf{Ace}) are in the lead, while \textsf{ADDMC}
significantly underperforms when combined with any of the previous encodings.
Our encoding \texttt{cw} significantly improves the performance of
\textsf{ADDMC}, making $\textsf{ADDMC}+\texttt{cw}$ comparable to
$\textsf{Ace}+\texttt{d02}$, $\textsf{c2d}+\texttt{bklm16}$, and
$\textsf{Cachet}+\texttt{sbk05}$. Furthermore,
\cref{tbl:tallies} shows that, while $\textsf{Ace}+\texttt{cd06}$ managed to
solve the most instances, $\textsf{ADDMC}+\texttt{cw}$ was the best-performing
algorithm-encoding combination on the largest number of instances. The scatter
plot on the left-hand side of \cref{fig:scatter} add to this by showing that
\texttt{cw} is particularly promising on Grid networks and tackles all DQMR
instances in less than a second. The scatter plot on the right-hand side of
\cref{fig:scatter} shows that \texttt{cw} is better than \texttt{sbk05} (i.e.,
the second-best encoding for \textsf{ADDMC}) on the majority of
instances. Seeing how, e.g., DQMR instances are trivial for
$\textsf{ADDMC}+\texttt{cw}$ but hard for $\textsf{Ace}+\texttt{cd06}$, and
vice versa for Mastermind instances, we conclude that the best-performing
algorithm-encoding combination depends significantly on (as-of-yet unknown)
properties of the Bayesian networks.

@@

We can explain what makes \textsf{ADDMC} run significantly faster with
\texttt{cw} than with any other encoding by considering asymptotic upper bounds
on the numbers of variables and ADDs based on the size and structure of the
Bayesian network. Let $n = |\mathcal{V}|$ be the number of vertices in the
Bayesian network, $d = \max_{X \in \mathcal{V}} |\mathrm{pa}(X)|$ the maximum
in-degree (i.e., the number of parents), and $v = \max_{X \in \mathcal{V}} |\im
X|$ the maximum number of values per variable. \Cref{tbl:asymptotes} shows how
\texttt{cw} has fewer variables and fewer ADDs than any other encoding. We
conjecture that it is primarily the reduced number of variables that makes the
\textsf{ADDMC} variable ordering heuristics much more effective. Note that these
are upper bounds and most encodings (including \texttt{cw}) can be smaller in
certain situations (e.g., with binary random variables or when a CPT has
repeating probabilities). We equate clauses and ADDs (more specifically, factors
of the function $\phi$ from \cref{alg:encoding}) here because \textsf{ADDMC}
interprets each clause of any WMC encoding as a multiplicative factor of the ADD
that represents the entire WMC instance \citep{DBLP:conf/aaai/DudekPV20}. For
literal-weight encodings, each weight is also a factor, but that does not affect
our asymptotic bounds.

\section{Conclusions and Future Work}
% TODO: maybe: emphasise that BN encodings have never been compared w.r.t. time
% before

@@

While we chose \textsf{ADDMC} as the WMC algorithm and Bayesian networks as a
canonical example of a probabilistic inference task, these are only examples
meant to illustrate the broader idea that choosing a more expressive
representation of weights can outperform increasing the size of the problem to
keep the weights simple. Indeed, in this work, we have provided a new
theoretical perspective on the expressive power of WMC and illustrated the
empirical benefits of that perspective. The same idea can be adapted to other
inference problem domains such as probabilistic programs
\citep{DBLP:journals/tplp/FierensBRSGTJR15,DBLP:journals/corr/abs-2005-09089} as
well as to search-based solvers such as \textsf{Cachet}
\citep{DBLP:conf/sat/SangBBKP04} and \textsf{DPMC}
---an extension to \textsf{ADDMC} that adds support for computations based on
tensors (rather than ADDs) and planning based on tree decompositions
\cite{DBLP:conf/cp/DudekPV20}. Another important direction for future work is to
develop a better understanding of what properties of Bayesian networks make an
instance easy for some algorithm-encoding combinations more than others.

>>>>

\documentclass[accepted]{uai2021}

\usepackage[british]{babel}
\usepackage{natbib} % has a nice set of citation styles and commands
\bibliographystyle{plainnat}
\renewcommand{\bibsection}{\subsubsection*{References}}
\usepackage{mathtools} % amsmath with fixes and additions
\usepackage[binary-units]{siunitx} % for proper typesetting of numbers and units
\usepackage{booktabs} % commands to create good-looking tables
\usepackage[table]{xcolor}
\usepackage{tikz} % nice language for creating drawings and diagrams

\usepackage{complexity}
\usepackage[capitalise]{cleveref}
\usepackage{multirow}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{subcaption}
\usepackage[ruled,vlined]{algorithm2e}

@@

\title{Weighted Model Counting with Conditional Weights for Bayesian Networks}
% TODO: maybe I should rotate the cumulative plot

% TODO: read 'Preparing the submission materials' (code: dilkas_167)
% TODO: update the abstract during submission

% The standard author block has changed for UAI 2021 to provide
% more space for long author lists and allow for complex affiliations
%
% All author information is authomatically removed by the class for the
% anonymous submission version of your paper, so you can already add your
% information below.

\author[1]{\href{mailto:Paulius Dilkas <p.dilkas@sms.ed.ac.uk>?Subject=Your UAI
    2021 paper}{Paulius~Dilkas}{}}
\author[1,2]{Vaishak~Belle}
\affil[1]{%
    School of Informatics\\
    University of Edinburgh\\
    Edinburgh, UK
}
\affil[2]{%
    Alan Turing Institute\\
    London, UK
}

@@

\begin{abstract}
  Weighted model counting (WMC) has emerged as the unifying inference mechanism
  across many (probabilistic) domains. Encoding an inference problem as an
  instance of WMC typically necessitates adding extra literals and clauses. This
  is partly so because the predominant definition of WMC assigns weights to
  models based on weights on literals, and this severely restricts what
  probability distributions can be represented. We develop a measure-theoretic
  perspective on WMC and propose a way to encode conditional weights on literals
  analogously to conditional probabilities. This representation can be as
  succinct as standard WMC with weights on literals but can also expand as
  needed to represent probability distributions with less structure. To
  demonstrate the performance benefits of conditional weights over the addition
  of extra literals, we develop a new WMC encoding for Bayesian networks and
  adapt a state-of-the-art WMC algorithm \textsf{ADDMC} to the new format. Our
  experiments show that the new encoding significantly improves the performance
  of the algorithm on most benchmark instances.
\end{abstract}

\section{Introduction}

@@

% TODO: must reference the SAT paper and explain the differences
% \cite{my_sat_paper}

@@

\begin{table*}
  \caption{Notation for a logic with two atoms. The elements in both columns are
    listed in the same order.}
  \label{tbl:notation_example}
  \centering
  \begin{tabular}{lcc}
    \toprule
    Name in logic & Boolean-algebraic notation & Set-theoretic notation \\
    \midrule
    Atoms (elements of $U$) & $a, b$ & $a, b$ \\
    \rowcolor{gray!10} Models (elements of $2^U$) & $\neg a \land \neg b, a \land \neg b, \neg a \land b, a \land b$ & $\emptyset, \{a\}, \{b\}, \{a, b\}$ \\
    & $\top$ & $\{ \emptyset, \{a\}, \{b\}, \{a, b\} \}$ \\
    & $\neg a \lor \neg b, a \to b$ & $\{ \emptyset, \{a\}, \{b\} \}, \{ \emptyset, \{b\}, \{a, b\} \}$ \\
    & $b \to a, a \lor b$ & $\{ \emptyset, \{a\}, \{a, b\} \}, \{ \{a\}, \{b\}, \{a, b\} \}$ \\
    & $\neg b, \neg a, a \leftrightarrow b$ & $\{\emptyset, \{a\}\}, \{\emptyset, \{b\}\}, \{\emptyset, \{a, b\}\}$ \\
    & $(a \land \neg b) \lor (b \land \neg a), a, b$ & $\{\{a\}, \{b\}\}, \{\{a\}, \{a, b\}\}, \{\{b\}, \{a, b\}\}$ \\
    & $\neg a \land \neg b, a \land \neg b, \neg a \land b, a \land b$ & $\{\emptyset\}, \{\{a\}\}, \{\{b\}\}, \{\{a, b\}\}$ \\
    \multirow{-7}{*}{Formulas (elements of $2^{2^U}$)} & $\bot$ & $\emptyset$ \\
    \bottomrule
  \end{tabular}
\end{table*}

@@

\subsection{Functions on Boolean Algebras}

@@

\begin{theorem} \label{prop:measure}
  The function $\mu_\nu$ is a measure.
\end{theorem}

@@

Since many measures of interest may not be factorable, a well-known way to
encode them into instances of WMC is by adding more literals
\citep{DBLP:journals/ai/ChaviraD08}. We can use the measure-theoretic
perspective on WMC to show that this is always possible, however, as ensuing
sections will demonstrate, it can make the inference task much harder in
practice.\footnote{The proofs of this and other theoretical results can be found
  in the supplementary material.}

\begin{theorem}
  For any set $U$ and measure $\mu\colon 2^{2^U} \to \mathbb{R}_{\ge 0}$, there
  exists a set $V \supseteq U$, a factorable measure $\mu'\colon 2^{2^V} \to
  \mathbb{R}_{\ge 0}$, and a formula $f \in 2^{2^V}$ such that $\mu(x) = \mu'(x
  \land f)$ for all formulas $x \in 2^{2^U}$.
\end{theorem}

@@

For a binary random variable $X$, $\mathrm{CPT}_X$ is simply a sum of smaller
functions, one for each row of the CPT. If $X$ has more than two values, we also
multiply $\mathrm{CPT}_X$ by `clause' functions that restrict the value of
$\phi(T)$ to zero whenever $|\mathcal{E}(X) \cap T| \ne 1$, i.e., we add mutual
exclusivity constraints that ensure that each random variable is associated with
exactly one value. Note that \citet{DBLP:conf/ijcai/ChaviraD07} use the same ADD
representation of CPTs for their compilation algorithm based on variable
elimination. For the example Bayesian network in \cref{fig:example_bn}, we get:
\begin{align*}
  \mathrm{CPT_F} &= 0.6[\lambda_{F=1}] \cdot [\lambda_{W=1}] + 0.4[\lambda_{F=0}] \cdot [\lambda_{W=1}] \\
                 &+ 0.1[\lambda_{F=1}] \cdot [\lambda_{W=0}] + 0.9[\lambda_{F=0}] \cdot [\lambda_{W=0}], \\
  \mathrm{CPT_T} &= ([\lambda_{T=l}] + [\lambda_{T=m}] + [\lambda_{T=h}]) \\
                 &\cdot (\overline{[\lambda_{T=l}]} + \overline{[\lambda_{T=m}]}) \cdot (\overline{[\lambda_{T=l}]} + \overline{[\lambda_{T=h}]}) \\
                 &\cdot (\overline{[\lambda_{T=m}]} + \overline{[\lambda_{T=h}]}) \cdot \dots.
\end{align*}

@@

\begin{lemma} \label{lemma:cpt}
  Let $X \in \mathcal{V}$ be a random variable with parents $\mathrm{pa}(X) = \{ Y_1,
  \dots, Y_n \}$. Then $\mathrm{CPT}_X\colon 2^{\mathcal{E}^*(X)} \to
  \mathbb{R}_{\ge 0}$ is such that for any $x \in \im X$ and $(y_1, \dots, y_n)
  \in \prod_{i=1}^n \im Y_i$,
  \[
    \mathrm{CPT}_X (T) = \Pr(X = x \mid Y_1 = y_1, \dots, Y_n = y_n),
  \]
  where $T = \{ \lambda_{X=x} \} \cup \{ \lambda_{Y_i=y_i} \mid i = 1, \dots, n
  \}$.
\end{lemma}

@@

\begin{lemma} \label{lemma:full_distribution}
  Let $\mathcal{V} = \{X_1, \dots, X_n\}$. Then
  \[
    \phi(T) =
    \begin{cases}
      \Pr(x_1, \dots, x_n) &
      \begin{aligned}
        &\text{if } T = \{ \lambda_{X_i=x_i} \}_{i = 1}^n \text{ for} \\
        &\text{some } \textstyle (x_i)_{i=1}^n \in \prod_{i=1}^n \im X_i
      \end{aligned} \\
      0 & \text{otherwise,}
    \end{cases}
  \]
  for all $T \in 2^U$.
\end{lemma}

We end with \cref{thm:correctness} that shows how $\phi$ can be combined with an
encoding of a single variable-value assignment so that \textsf{ADDMC}
\citep{DBLP:conf/aaai/DudekPV20} would compute its marginal probability.

\begin{theorem} \label{thm:correctness}
  For any $X \in \mathcal{V}$ and $x \in \im X$,
  \[
    (\exists_U(\phi \cdot [\lambda_{X=x}]))(\emptyset) = \Pr(X = x).
  \]
\end{theorem}

@@

We compare the six WMC encodings for Bayesian networks when run with both
\textsf{ADDMC} \citep{DBLP:conf/aaai/DudekPV20} and the WMC algorithms used in
the original papers.\footnote{Both \texttt{cd05} and \texttt{cd06} cannot be run
  with most WMC algorithms including \textsf{ADDMC} because these encodings
  allow for additional models that the WMC algorithm is supposed to ignore
  \citep{DBLP:conf/ijcai/ChaviraD05,DBLP:conf/sat/ChaviraD06}.} We compare the
encodings with respect to the total time it takes to encode a Bayesian network,
compile it or run a WMC algorithm on it, and extract the (numerical) answer.
Note that while all five papers that introduce other encodings include
experimental comparisons of encoding size, that is not feasible with
\textsf{ADDMC} as even instances that are fully solved in less than
\SI{0.1}{\second} are too big to build the full ADD within reasonable time and
memory limits. The experiments were run on a computing cluster with Intel Xeon
Gold 6138 and Intel Xeon E5-2630 processors\footnote{Each instance is run on the
  same processor for all encodings.} running Scientific~Linux~7 with a
\SI{32}{\gibi\byte} memory limit and a \SI{1000}{\second} timeout on both
encoding and inference. For inference, we use \textsf{Ace} for \texttt{cd05}
\citep{DBLP:conf/ijcai/ChaviraD05}, \texttt{cd06}
\citep{DBLP:conf/sat/ChaviraD06}, and \texttt{d02}
\citep{DBLP:conf/kr/Darwiche02};
\textsf{Cachet}\footnote{\url{https://cs.rochester.edu/u/kautz/Cachet/}}
\citep{DBLP:conf/sat/SangBBKP04} for \texttt{sbk05}
\citep{DBLP:conf/aaai/SangBK05}; and \textsf{c2d}
\citep{DBLP:conf/ecai/Darwiche04} for compilation and \textsf{query-dnnf}
\footnote{\url{http://www.cril.univ-artois.fr/kc/d-DNNF-reasoner.html}} for
answer computation for \texttt{bklm16} \citep{DBLP:conf/ecai/BartKLM16}. For
encoding, we use \textsf{bn2cnf}
\footnote{\url{http://www.cril.univ-artois.fr/KC/bn2cnf.html}} for
\texttt{bklm16}, and \textsf{Ace} for all other encodings (except for
\texttt{cw}, which is implemented in Python).

\textsf{Ace} was not used to encode evidence, as preliminary experiments
revealed that the evidence-encoding implementation contains bugs that can lead
to incorrect answers or a Java exception being thrown on some instances of the
data set (and the source code is not publicly available). Instead, we simply
list all the evidence as additional clauses in the encoding. Furthermore, to
ensure that \texttt{bklm16} \citep{DBLP:conf/ecai/BartKLM16} (whether run with
\textsf{ADDMC} \citep{DBLP:conf/aaai/DudekPV20} or \textsf{c2d}
\citep{DBLP:conf/ecai/Darwiche04}) returns correct answers on most instances, we
had to disable one of the improvements that \texttt{bklm16} brings over
\texttt{cd06} \citep{DBLP:conf/sat/ChaviraD06}, namely, the construction of a
scaling factor that `absorbs' one probability from each CDT
\citep{DBLP:conf/ecai/BartKLM16}. For realistic benchmark instances, this
scaling factor can easily be below $10^{-30}$, and thus would require
arbitrary-precision floating-point arithmetic to be usable. Even a toy Bayesian
network with seven binary independent variables with probabilities $0.1$ and
$0.9$ is enough for \textsf{bn2cnf} to output precisely zero as the scaling
factor. We note that this issue likely remained unnoticed because
\citet{DBLP:conf/ecai/BartKLM16} did not attempt to compute numerical answers in
their experiments.

@@

For experimental data, we use the Bayesian networks available with \textsf{Ace}
and \textsf{Cachet} \citep{DBLP:conf/sat/SangBBKP04}, most of which happen to be
binary. We classify them into the following seven categories:
\begin{itemize*}
\item DQMR and
\item Grid networks as described by \citet{DBLP:conf/aaai/SangBK05},
\item Mastermind, and
\item Random Blocks from the work of \citet{DBLP:journals/ijar/ChaviraDJ06},
\item remaining binary Bayesian networks that include Plan Recognition
  \citep{DBLP:conf/aaai/SangBK05}, Friends and Smokers, Students and Professors
  \citep{DBLP:journals/ijar/ChaviraDJ06}, and \texttt{tcc4f}, and
\item non-binary classic Bayesian networks (\texttt{alarm}, \texttt{diabetes},
  \texttt{hailfinder}, \texttt{mildew}, \texttt{munin1}--\texttt{4},
  \texttt{pathfinder}, \texttt{pigs}, \texttt{water}).
\end{itemize*}

\Cref{fig:cumulative} shows that \texttt{cd05}
\citep{DBLP:conf/ijcai/ChaviraD05} and \texttt{cd06}
\citep{DBLP:conf/sat/ChaviraD06} (when run with \textsf{Ace}) are in the lead,
while \textsf{ADDMC} \citep{DBLP:conf/aaai/DudekPV20} significantly
underperforms when combined with any of the previous encodings. Our encoding
\texttt{cw} significantly improves the performance of \textsf{ADDMC}, making
$\textsf{ADDMC}+\texttt{cw}$ comparable to $\textsf{Ace}+\texttt{d02}$,
$\textsf{c2d}+\texttt{bklm16}$, and $\textsf{Cachet}+\texttt{sbk05}$.
Furthermore, \cref{tbl:tallies} shows that, while $\textsf{Ace}+\texttt{cd06}$
managed to solve the most instances, $\textsf{ADDMC}+\texttt{cw}$ was the
best-performing algorithm-encoding combination on the largest number of
instances. The scatter plot on the left-hand side of \cref{fig:scatter} add to
this by showing that \texttt{cw} is particularly promising on Grid networks and
tackles all DQMR instances in less than a second. The scatter plot on the
right-hand side of \cref{fig:scatter} shows that \texttt{cw} is better than
\texttt{sbk05} \citep{DBLP:conf/aaai/SangBK05} (i.e., the second-best encoding
for \textsf{ADDMC}) on the majority of instances. Seeing how, e.g., DQMR
instances are trivial for $\textsf{ADDMC}+\texttt{cw}$ but hard for
$\textsf{Ace}+\texttt{cd06}$, and vice versa for Mastermind instances, we
conclude that the best-performing algorithm-encoding combination depends
significantly on (as-of-yet unknown) properties of the Bayesian networks.

@@

We can explain what makes \textsf{ADDMC} \citep{DBLP:conf/aaai/DudekPV20} run
significantly faster with \texttt{cw} than with any other encoding by
considering asymptotic upper bounds on the numbers of variables and ADDs based
on the size and structure of the Bayesian network. Let $n = |\mathcal{V}|$ be
the number of vertices in the Bayesian network, $d = \max_{X \in \mathcal{V}}
|\mathrm{pa}(X)|$ the maximum in-degree (i.e., the number of parents), and $v =
\max_{X \in \mathcal{V}} |\im X|$ the maximum number of values per variable.
\Cref{tbl:asymptotes} shows how \texttt{cw} has fewer variables and fewer ADDs
than any other encoding. We conjecture that it is primarily the reduced number
of variables that makes the \textsf{ADDMC} variable ordering heuristics much
more effective. Note that these are upper bounds and most encodings (including
\texttt{cw}) can be smaller in certain situations (e.g., with binary random
variables or when a CPT has repeating probabilities). We equate clauses and ADDs
(more specifically, factors of the function $\phi$ from \cref{alg:encoding})
here because \textsf{ADDMC} interprets each clause of any WMC encoding as a
multiplicative factor of the ADD that represents the entire WMC instance
\citep{DBLP:conf/aaai/DudekPV20}. For literal-weight encodings, each weight is
also a factor, but that does not affect our asymptotic bounds.

\section{Conclusions and Future Work}

@@

While we chose \textsf{ADDMC} \citep{DBLP:conf/aaai/DudekPV20} as the WMC
algorithm and Bayesian networks as a canonical example of a probabilistic
inference task, these are only examples meant to illustrate the broader idea
that choosing a more expressive representation of weights can outperform
increasing the size of the problem to keep the weights simple. Indeed, in this
work, we have provided a new theoretical perspective on the expressive power of
WMC and illustrated the empirical benefits of that perspective. The same idea
can be adapted to other inference problem domains such as probabilistic programs
\citep{DBLP:journals/tplp/FierensBRSGTJR15,DBLP:journals/corr/abs-2005-09089} as
well as to search-based solvers such as \textsf{Cachet}
\citep{DBLP:conf/sat/SangBBKP04} and \textsf{DPMC} ---an extension to
\textsf{ADDMC} that adds support for computations based on tensors (rather than
ADDs) and planning based on tree decompositions \cite{DBLP:conf/cp/DudekPV20}.
Another important direction for future work is to develop a better understanding
of what properties of Bayesian networks make an instance easy for some
algorithm-encoding combinations more than others.
